### 2013-10-25 src2/thesis.tex
update

\begin{enumerate}[A.] 
\item Josef Svenningsson and Bo Joel Svensson. Simple and Compositional Reification of Monadic Embedded Languages, 2013. \emph{18th ACM SIGPLAN International Conference of Functional Programming, ICFP 2013.}
\item Bo Joel Svensson, Mary Sheeran and Koen Claessen. Obsidian: A Domain Specific Embedded Language for General Purpose Parallel Programming of Graphics Processors. In \emph{Proc. of Implementation and Applications of Functional Languages (IFL)}, Lecture Notes in Computer Science, Springer Verlag, March 2009.
\item Bo Joel Svensson, Koen Claessen and Mary Sheeran. GPGPU kernel implementation and refinement using Obsidian. \emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
\item Koen Claessen, Mary Sheeran and Bo Joel Svensson. Expressive Array Constructs in an Embedded GPU Kernel Programming Language. In \emph{Proceedings of the 7th workshop of Declarative aspects and applications of multicore programming}, DAMP '12. ACM. 
\item Josef Svenningsson, Bo Joel Svensson and Mary Sheeran. Efficient Counting Sort Implementations using an Embedded GPU Programming Language. \emph{2nd Workshop of Functional High-Performance Computing, FHPC '13.}
\item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
\item Bo Joel Svensson and Ryan Newton. Programming Future Parallel Architectures with Haskell and ArBB. \emph{Future Architectural Support for Parallel Programming (FASPP), in conjunction with ISCA '11.}
\item Bo Joel Svensson and Mary Sheeran. Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks. \emph{Proceedings of the 1st ACM SIGPLAN Workshop of Functional High-Performance computing, FHPC '12.}
\end{enumerate}

@@

The code and examples in section one and two are my work. Section four we developed together 
but is mostly Josef's expertise. For section four we also got very much and valuable help 
from Emil Axelsson.

@@

\subsubsection{Papers G and H} 
Here, again, most implementation effort was done by me. For paper G, this work was performed 
under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
paper. Programming and running time experimentation was performed by me, while getting much 
valuable advice and guidance from Ryan.

@@

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possibility for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still making building blocks implemented by the programmer reusable and composable. 
We want to give more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian.

@@

A CUDA GPU is built around a single kind of processor (opposed to the different 
kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
MultiProcessors), all contain a number of function units called SPs (Streaming Processors). 
Each MP also contains local memory, called shared memory since it can be accessed 
by all of the SPs in that MP. The number of MPs vary over the available GPUs; cheaper 
GPUs have as few as one MP, and as you go up in price the number of MPs increase.

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
a {\em Warp}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
local shared memory. The maximum block size is currently 1024 threads. Since 
the number of threads in a block can be higher than the warp size, a barrier 
synchronisation mechanism exists to ensure that threads within a block 
can communicate safely via the shared memory.

Papers B, C, E and F (sections~\ref{sec:paperB},~\ref{sec:paperC},~\ref{sec:paperE},
~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
about them ,in relation to Obsdian.

@@

The CUDA programming model reflects the hierarchical architecture of GPUs. The programmer 
writes what is called {\em kernels}, in CUDA jargon a kernel refers to a sequential C program 
that is parameterised on a threads identity. The kernel program is executed in parallel across 
all the threads of a block and many such blocks of threads can execute in parallel on the 
available MPs. The programmer needs to decompose the algorithm and data in such a way what 
communication will be local to the blocks.

@@

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in a host language, Haskell in this case. 
The embedded language approach is beneficial since we are not entirely sure how 
to program current and future highly parallel computers. Being embedded allows 
us more rapid prototyping of ideas. Other benefits of being embedded often listed 
are: making use of the host language's type system, parser and libraries (to some extent) 
and to have the host language as a powerful macro language for your embedded language.

A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embedd a compiler 
backend and generate efficient code in some target language. One key aspect of this 
approach is that the embedded language is a library of functions that create and 
compose Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
embedding that builds ASTs is called a deep embedding. Shallow embeddings on the 
other hand do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
Josef Svenningsson and Emil Axelsson combine the two methods, shallow and deep, 
in order to simplify the AST data type (fewer constructors) and make more extensible 
embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings, while EmbArBB is a more traditional deep embedding.

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, there is Lava and Wired in the fields of hardware design and 
verification and Feldspar for digital signal processing~\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
programming there are numerous embedded approaches using various host languages. 
But to mention two that use Haskell we have Accelerate and Nikola 
\citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
goal of being parallel across platforms \citet{ARBB2011}.
Related work and information about the tools, libraries and language that make the 
wider context surrounding Obsidian is described in section~\ref{sec:relatedwork}.

@@

Obsidian aspires to raise the level of abstraction of GPU kernel implementation 
while maintaining enough control over details that influence performance on a GPU. 
We try to reach these goals by offering abstractions to the programmer that are 
compositional and parallel. We also ensure that these can easily be compiled 
to NVIDIA CUDA code. We have two different array representations, each with 
its strengths and weaknesses. We also feel that it is important that the programmer 
be able to use local shared memory in computations and has the ability to influence 
memory access patterns. What these array representations and abstractions are 
will be explained in the following sections, with references to the papers 
that introduce them or explain them the best.

@@

The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
in the hierarchy leading to warp programming being a special case. One benefit a warp has is 
that threads can communicate via shared memory without using synchronisation primitives. There 
is no programmer support for warp level programming in CUDA (as there is for block level 
programs). The programmer needs to take care to ensure that the communication patterns are 
within a warp and is then free to remove synchronisations. If warps where to be put in the 
program level hierarchy of Obsidian, its place would be between the thread and block level. 
This forces the programmer to go via warps when putting together a block computation even 
if the communication pattern is not suited for that, leading to inconvenience with no gain.

As an example of how the hierarchy level parameter is used I show the {\tt ForAll} constructor 
from the {\tt Program} data type.  This constructor represents parallelism either over threads 
or blocks.

@@

The approach taken to embedd the scalar language is very similar to what is described 
in~\citet{COMPILEEDSL}; but we use a Generalised Algebraic Datatype (GADT) to obtained 
typed expressions rather than phantom types.

@@

Push arrays where added to Obsidian as a complement to pull arrays. Pull arrays are 
good because they are so simple to understand and easy to parallelise. However certain
operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
or interleaving of pull arrays result in conditionals in the indexing function. The 
interleaving case, which is worst, leads to diverging branching in every warp. If branches 
diverge the GPU will execute turn off the processing elements taking one path and 
allow the others to progress; then turning to the other branch. The computation of the 
branches has become serialised and compute resources are wasted.

>>>>

\begin{enumerate}[A.] 
\item Josef Svenningsson and Bo Joel Svensson. Simple and Compositional Reification of Monadic Embedded Languages, 2013. \emph{18th ACM SIGPLAN International Conference of Functional Programming, ICFP 2013.}
\item Bo Joel Svensson, Mary Sheeran and Koen Claessen. Obsidian: A Domain Specific Embedded Language for General Purpose Parallel Programming of Graphics Processors. In \emph{Proc. of Implementation and Applications of Functional Languages (IFL)}, Lecture Notes in Computer Science, Springer Verlag, March 2009.
\item Bo Joel Svensson, Koen Claessen and Mary Sheeran. GPGPU kernel implementation and refinement using Obsidian. \emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
\item Koen Claessen, Mary Sheeran and Bo Joel Svensson. Expressive Array Constructs in an Embedded GPU Kernel Programming Language. In \emph{Proceedings of the 7th workshop on Declarative aspects and applications of multicore programming}, DAMP '12. ACM. 
\item Josef Svenningsson, Bo Joel Svensson and Mary Sheeran. Efficient Counting Sort Implementations using an Embedded GPU Programming Language. \emph{2nd Workshop on Functional High-Performance Computing, FHPC '13.}
\item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
\item Bo Joel Svensson and Ryan Newton. Programming Future Parallel Architectures with Haskell and ArBB. \emph{Future Architectural Support for Parallel Programming (FASPP), in conjunction with ISCA '11.}
\item Bo Joel Svensson and Mary Sheeran. Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks. \emph{Proceedings of the 1st ACM SIGPLAN Workshop on Functional High-Performance computing, FHPC '12.}
\end{enumerate}

@@

The code and examples in section one and two are my work. The contents of section four was 
developed by Josef Svenningsson and me. For section four we also got very much and valuable help 
from Emil Axelsson.

@@

\subsubsection{Papers G and H} 
The work behind paper G, work was performed by me 
under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
paper. Programming and running time experimentation was performed by me, while getting much 
valuable advice and guidance from Ryan.

@@

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possibility for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still making building blocks implemented by the programmer reusable and composable. 
We want to provide more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian.

@@

A CUDA GPU is built around a single kind of processor (as opposed to the different 
kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
MultiProcessors) all contain a number of function units called SPs (Streaming Processors). 
Each MP also contains local memory, called shared memory since it can be accessed 
by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
GPUs have as few as one MP, and as you go up in price the number of MPs increases.

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
local shared memory. The maximum block size is currently 1024 threads. Since 
the number of threads in a block can be higher than the warp size, a barrier 
synchronisation mechanism exists to ensure that threads within a block 
can communicate safely via the shared memory.

Papers B, C, E and F (sections~\ref{sec:paperB},~\ref{sec:paperC},~\ref{sec:paperE},
~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
about them in relation to Obsdian.

@@

The CUDA programming model reflects the hierarchical architecture of GPUs. The programmer 
writes what are called {\em kernels}; in CUDA jargon a kernel refers to a sequential C program 
that is parameterised on a thread's identity. The kernel program is executed in parallel across 
all the threads of a block and many such blocks of threads can execute in parallel on the 
available MPs. The programmer needs to decompose the algorithm and data in such a way that 
communication will be local to the blocks.

@@

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in a host language, Haskell in this case. 
The embedded language approach is beneficial since we are not entirely sure how 
to program current and future highly parallel computers. Implementing embedded languages allows 
us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
are: making use of the host language's type system for guaranteeing generating correctly typed 
target code, parser and libraries (to some extent) 
and to have the host language as a powerful macro language for your embedded language.

A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embed a compiler 
backend and generate efficient code in some target language. One key aspect of this 
approach is that the embedded language is a library of functions that create and 
compose Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
embedding that builds ASTs is called a deep embedding. Shallow embeddings, on the 
other hand, do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
Josef Svenningsson and Emil Axelsson combine the two methods, shallow and deep, 
in order to simplify the AST data type (fewer constructors) and make more extensible 
embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings, while EmbArBB is a more traditional deep embedding.

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, Lava and Wired in the fields of hardware design and 
verification and Feldspar for digital signal processing~\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
programming, there are numerous embedded approaches using various host languages. 
Two languages for GPU programming embedded in Haskell are  Accelerate and Nikola 
\citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
goal of being parallel across platforms \citet{ARBB2011}.
Related work and information about the tools, libraries and language that make the 
wider context surrounding Obsidian can be found in section~\ref{sec:relatedwork}.

@@

Obsidian aspires to raise the level of abstraction of GPU kernel implementation, 
while maintaining enough control over details that influence performance on a GPU. 
We try to reach these goals by offering abstractions to the programmer that are 
compositional and parallel. We also ensure that these can easily be compiled 
to NVIDIA CUDA code. We have two different array representations, each with 
its strengths and weaknesses. We also feel that it is important that the programmer 
be able to use local shared memory in computations and have the ability to influence 
memory access patterns. What these array representations and abstractions are 
will be explained in the following sections, with references to the papers 
that introduce them or explain them the best.

@@

The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
in the hierarchy leading to warp programming being a special case. One benefit a warp has is 
that threads can communicate via shared memory without using synchronisation primitives. There 
is no programmer support for warp level programming in CUDA (as there is for block level 
programs). The programmer needs to take care to ensure that the communication patterns are 
within a warp and is then free to remove synchronisations. If warps were to be put in the 
program level hierarchy of Obsidian, its place would be between the thread and block level. 
This would force the programmer to go via warps when putting together a block computation even 
if the communication pattern is not suited for that, leading to inconvenience with no gain.

As an example of how the hierarchy level parameter is used, I show the {\tt ForAll} constructor 
from the {\tt Program} data type.  This constructor represents parallelism either over threads 
or blocks.

@@

The approach taken to embed the scalar language is very similar to what is described 
in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
Datatype (GADT) to obtain typed expressions. An alternative to the GADT, is to use phantom 
types. Earlier versions of Obsidian (section~\ref{sec:paperB} used phantom types.

@@

Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
good because they are so simple to understand and easy to parallelise. However certain
operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
or interleaving of pull arrays result in conditionals in the indexing function. The 
interleaving case, which is worst, leads to diverging branching in every warp. If branches 
diverge the GPU will execute turn off the processing elements taking one path and 
allow the others to progress; then turning to the other branch. The computation of the 
branches has become serialised and compute resources are wasted.
